---
title: "MB5370 Module 04. Workshop 1 - Introduction"
author: "Ben Cresswell"
output: html_document
date: "2024-03-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




# Packages
```{r}
library(tidyverse)
# source("/code/packages.r")
library(ggplot2)
```


# Load the data (mpg is built into ggplot2)
```{r}
data(mpg)
```


# Quick data checks
```{r}
head(mpg)
glimpse(mpg)
summary(mpg)
```


# Create first plot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

The plot shows a negative relationship between engine size (displ) and fuel efficiency (hwy). In other words, cars with big engines use more fuel. What does this say about fuel efficiency and engine size?



# Understanding grammar of graphics
What happens if we just try to run ggplot on its own?
```{r}
ggplot() 
# Or with just the data specified?
ggplot(data = mpg)
```

We need to map some aesthetics!!

# When you’re creating a plot, you essentially need two attributes of a plot: a geom and aesthetics.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```


# Change point colour by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

# Change point size by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```
Note the warning!!!


# Change transparency (alpha) by class
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```
Another warning!!
Question: When would using alpha (or size be appropriate??)

# Change point shape by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```


# Make all points blue
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```


# Question 1:
#### What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(colour = displ < 5), color = "blue")
```

# Answer:
#### The function aes() requires x and y to be specified. In this case an error occured and the chunk could not be run


# Troubleshooting 

## Example of a common mistake

#### ggplot(data = mpg) 
#### + geom_point(mapping = aes(x = displ, y = hwy)) 
#### # the + should be on top line


# Faceting

## facet_wrap
Split the plot out by car type (class)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```


## facet_grid
A separate facet for each combination of drive-type (e.g. 4WD) * number of cylinders
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```
Note that there are no occurrences of 5 cylinder 4WDs OR RWD vehicles!


## facet_grid by just row (or column)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ drv)
  #facet_grid(. ~ cyl) # Alternatively

```

# Exercise:
Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? 

```{r}
?facet_wrap
```

#### nrow and ncol determine the number of rows and column in the grid, respectively.
#### dir is another option that permit to control the layout of the panels

# Lines
We already used points:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```


However, ggplot2 can use a variety of geom objects to represent the data. Here, we might want to use bar plots, line charts, boxplots and so on. Well we can handle this issue in ggplot directly using a different geom to plot the same data. Here, instead of plotting points, we will use a smooth line. 

# To display the same data as a smooth line fit through the points use geom_smooth().
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
### Question: whey don't we use geom_line() here? What would that look like? ###

#### Using the function geom_line() will produce a straight line that follows the mean values between the points fluctuation, without ahowing the real value trend.

So let’s recap. A geom is an object that your plot uses to represent the data. To change the geom type in your plot, simply change the geom function that you add to your plot template. Sometimes you may want to try a few things out, in which case you could use comments to help you remember what worked and what didn’t. 

## Using comments (#)
```{r}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy)) + # points horrible 
  geom_smooth(mapping = aes(x = displ, y = hwy)) # try smooth line
```

Question: how does R work out where to plot the line??? Can use the chunk output to inform us. Can also use the help menu.

# Changing linetype
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) # Can also use "lty = " for the same result
```


# Grouping
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```
 
# Change line colour based on drv value
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = T,
  )
```

# Multiple geoms
We already did this one way
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

A better way...
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```


# Can still manipulate each geom/layer separately:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

Now we have different colours to indicate the raw data (geom_point) and one colour for the overall relationship (geom_smooth). Neat!


```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = mpg %>% filter(class == "subcompact"), se = T)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = F)
```


Question: what does the "se = FALSE" do?
### It visualizes the standard error around our line

# Exercise:
1.  What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

#### geom_line(), geom_boxplot, geom_hist(), geom_area()

2.  Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

```{r}
ggplot(data = mpg, aes(x=displ, y=hwy)) +
  geom_line(mapping = aes(color = class)) 
  
ggplot(data = mpg, aes(x=displ, y=hwy)) +
  geom_boxplot(mapping = aes(color = class)) 

ggplot(data = mpg, aes(x=displ)) +
  geom_histogram(mapping = aes(color = class)) 
  
ggplot(data = mpg, aes(x=displ, y=hwy)) +
  geom_area(mapping = aes(color = class)) 
  
  
```


3.  Will these two graphs look different? Why/why not?
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

#### They will look the same, since the function is the same, just constructed 

# Transformations and stats

## Load data
```{r}
data("diamonds")
glimpse(diamonds)
```

# Bar plot
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

# Overriding defaults
Make some new data
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo
```


```{r}
ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```
Question: Why did we specify now specify a y axis?

#### To override the default stat

# What if we want to know the proportion rather than the absolute frequency?
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
```

Question: does anyone get the warning that's in the workbook?
Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
If so, can use # geom_bar(mapping = aes(x = cut, y = stage(after_stat = prop), group = 1))

# Plotting statistical details
```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```


# Aesthetic adjustments adjustments
Another way to boost the way you can convey information with plots using ggplot2 is to use aesthetics like colour or fill to change aspects of bar colours. We already did this once, but there are multiple options available to you, including changing the fill our outline colours.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```
Question: Does anyone notice anything different in the colour scheme of this plot?
(Hint: It's in the viridis colour palette (colour blind friendly), but why is it different from the colour palette we used earlier?)

#### BEcause specifying "fill", it fills the whole bars

Check out the difference:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) 

ggplot(data = diamonds, mapping = aes(x = color, y = depth)) + 
  geom_point(mapping = aes(color = cut)) 
```


# Filling by a variable
Now try using these aesthetics to colour by another variable like clarity. Notice how the stacking is done automatically. This is done behind the scenes with a position argument.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```


# To alter transparency (alpha)
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
```


# To color the bar outlines with no fill color
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```


# Position adjustments
position = "fill" works like stacking, but makes each set of stacked bars the same height.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```


# position = "dodge" 
Places overlapping objects directly beside one another.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

# Jittering
position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.
```{r}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
  geom_point(mapping = aes(x = displ, y = hwy))
```





# Assignment 

```{r}
df_snake <- wpd_datasets

ggplot(data = df_snake, aes(x=X, y=Y)) +
  geom_line()
```

# 3.1 Labels

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se.e = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")

```

## Adding subtitle, caption and labels

```{r}

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov",
     x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type"
  )


```


# 3.2 Annotations

```{r}
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class)

```

# 3.3 Scales

## R does it in automatic, but we can offer values to the scale parameters

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete()


```


# 3.4 Axis ticks

## Modify axis ticks

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))
```

```{r}
seq(15, 40, by = 5)
```
## Seq produce a sequence of number based on the limit and sequence parameter we provide

## Suppress the labels altogether

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL)
```

# 3.5 Legends and colour schemes 

## Change legend position

```{r}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right") # the default
```

## To suppress the legend --> legend.position = "none"

# 3.6 Replacing a scale 

## Scaling in Log10()

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10()
```

## Colour scale

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1")
```

## Redundant shape mapping

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_colour_brewer(palette = "Set1")


```

## Set the colour you prefer

```{r}
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue"))


```


## Hottest colour packages: viridis!

```{r}
install.packages('viridis')
install.packages('hexbin')
library(viridis)
library(hexbin)
```

## Trying out viridis

```{r}
df <- tibble( # note we're just making a fake dataset so we can plot it
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + # a new geom!
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()

```

# 3.7 Themes

## Trying the themes of ggplot 2

```{r}

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()


ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light()


ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic()


ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark()


```

## Example of a personalized theme

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Arial")) 
```

# 3.8 Saving and exporting your plots

## Saving the most recent plot with ggsave()

```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point()

ggsave("my-plot.pdf")
#> Saving 7 x 4.32 in image
```

## We can adjust the dimensions of the plot with the "width" and "height" arguments

# 3.10 Assignment: End-to-end data analysis in R

## Importing the dataframe

```{r}
install.packages("janitor")
library("janitor")

data_frame <- read.csv("../data/export.csv") |> 
  row_to_names(row_number = 1) |>
  clean_names()
head(data_frame)
```

## Clean the dataset removing missing values, variable of no interest and extracting subset of different fishing mode

```{r}
combination <- data_frame |>
  select(x | contains("Combination")) |>  
  row_to_names(row_number = 1) |>
  mutate(across(is.character, as.numeric)) |>
  filter(!is.na(CalendarYear))
  
diving <- data_frame |>
  select(x | contains("diving")) |>  
  row_to_names(row_number = 1) |>
  mutate(across(is.character, as.numeric)) |>
  filter(!is.na(CalendarYear))

line <- data_frame |>
  select(x | contains("line")) |>  
  row_to_names(row_number = 1) |>
  mutate(across(is.character, as.numeric)) |>
  filter(!is.na(CalendarYear))

other <- data_frame |>
  select(x | contains("other")) |>  
  row_to_names(row_number = 1) |>
  mutate(across(is.character, as.numeric)) |>
  filter(!is.na(CalendarYear))

net <- data_frame |>
  select(x | contains("net")) |>  
  row_to_names(row_number = 1) |>
  mutate(across(is.character, as.numeric)) |>
  filter(!is.na(CalendarYear))
```

```{r}
comb_clean <- subset(combination, select = -c(Tonnes, DiscardNumber))
comb_clean <- na.omit(comb_clean)

diving_clean <- subset(diving, select = -c(Tonnes, DiscardNumber))
diving_clean <- na.omit(diving_clean)

line_clean <- subset(line, select = -c(Tonnes, DiscardNumber))
line_clean <- na.omit(line_clean)

other_clean <- subset(other, select = -c(Tonnes, DiscardNumber))
other_clean <- na.omit(other_clean)

net_clean <- subset(net, select = -c(Tonnes, DiscardNumber))
net_clean <- na.omit(net_clean)
```


## If you want wide data do this
## Create wide-format versions of each cleaned dataset by renaming the columns to specific fishing method

```{r}
comb_clean_wide <- comb_clean |>
  rename(comb_licence = Licences, 
         comb_days = Days)

line_clean_wide <- line_clean|>
  rename(line_licence = Licences, 
         line_days = Days)

other_clean_wide <- other_clean|>
  rename(other_licence = Licences, 
         other_days = Days)

diving_clean_wide <- diving_clean|>
  rename(diving_licence = Licences, 
         diving_days = Days)
```

## Join single dataset together, searching and joining columns with shared names

```{r}
join_data <- comb_clean |>
  full_join(line_clean) |>
  full_join(other_clean) |>
  full_join(diving_clean)
```
## If you want  long data do this (hint - do this)
## Add a new column "Fish_type" to each single dataset with the same value for each row (same fishing method in the single dataset)

```{r}
comb_clean_long <- comb_clean |>
  mutate(Fish_type = "Combined")

line_clean_long <- line_clean|>
  mutate(Fish_type = "Line")

other_clean_long <- other_clean|>
  mutate(Fish_type = "Other")

diving_clean_long <- diving_clean|>
  mutate(Fish_type = "Diving")
```

## Combine all four datasets into one long-format dataset.
## The command bind_rows() stacks the rows from all the datasets vertically.

```{r}
all_data <- bind_rows(comb_clean_long, line_clean_long, other_clean_long, diving_clean_long)


```
 
## Create a graph to visualize the licences per year for each fishing method


```{r}
all_data$CalendarYear <- as.numeric(as.character(all_data$CalendarYear))
ggplot(all_data, aes(x = CalendarYear, y = Licences, color = Fish_type)) +
  geom_line(size = 1) +
  scale_x_continuous(breaks = pretty(all_data$CalendarYear)) + 
  theme_minimal() +
  labs(title = "LIcences per year per fish type",
       y = "Licences number",
       x = "Year")
```
 
## This line graph shows the number of licences issued each year, grouped by fishing method (Fish_type). 
## Trends over time are illustrated for each method: Combined, Line, Diving, and Other.
## This allows for comparison in licence activity across different fishery types over time.

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

long_data <- all_data |>
  pivot_longer(cols = c(Licences, Days), names_to = "Variable", values_to = "Value") 

# Scale the days of linefishing because too high and prevent the other values to stand out in the graph


# Plot both Licences and Days over CalendarYear
ggplot(long_data, aes(x = CalendarYear, y = Value, color = Variable, linetype = Variable)) +
  geom_line(size = 1) +
  facet_wrap(~Fish_type, scales = "free") +  # Optional: to separate by fish type
  theme_minimal() +
  labs(title = "Licences and Days per Year",
       x = "Year",
       y = "Value")
```

## This graph shows trends over time for both the number of licences and fishing days, separated by fishing method.
## Each panel (facet) represents a different fishing type, with lines distinguishing between licences and days. 
## This layout allows easy comparison of fishing effort and activity across methods and years.
## It's interesting to notice how different fishing methods show a sharp rise in days in different periods, mostly corresponding to a less sharp ## amount in licences for the same years.







# Workshop 3 - Data wrangling in R

# 4.4.1 Excercise

```{r}
library(tidyverse)
``


## Computing rate in table 1

```{r}
table1 |>
  mutate(rate = cases / population * 10000)

```

## Computing per year

```{r}
table1 %>% 
  count(year, wt = cases)

```

## Let's see changes over time

```{r}
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

```


# 1.
## Table 1
#### First column represent the country where the data were taken; 
#### Second column shows the years; 
#### Third column shows the number of cases of an occurrence within the population; 
#### Fourth column represent the population count of the country in that year; 
#### Fifth column represents the ratio between the cases and the entire population 

##Table 2
#### Column 1: Country
#### Column 2: year
#### Column 3: Type of count
#### Cloumn 4: Count of the cases or population

## Table 3
#### Column 1: Country
#### Column 2: year
#### Column 3: rate between casea and population

# 2.
## a.
#### Create a dataset that contains count specifying each country and year where "type" is "cases"
#### Repeat specifying "population" as a type


## pivot_longer() in Billboard dataset

```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )

```

## Dropping the NA values will make the dataset shorter

```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )

```

## 4.5.2

```{r}
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)


```


## Dataset with three variables

```{r}
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )

```

## 4.5.3


```{r}
cms_patient_experience
```

```{r}
cms_patient_experience |> 
  distinct(measure_cd, measure_title)

```

## pivot_wider() needs existing column that specify the values and the names

```{r}
cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )

```

## Still organization spread in different rows. Bring them together with pivot_wider()

```{r}
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )

```


## 4.5.4

## Tribble with blood pressure of two patients

```{r}
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)


```

```{r}
df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )

```

```{r}
df |> 
  distinct(measurement) |> 
  pull()

```

```{r}
df |> 
  select(-measurement, -value) |> 
  distinct()


```

## plot_wider combines the values to create an empty dataframe

```{r}
df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)

```

## 4.5.5 Exercise

## 4.5.6 Separating and uniting data tables

##  Separating a column in two values
```{r}
table3

table3 %>% 
  separate(rate, into = c("cases", "population"))

```

